Characterization of transient resonances in extreme-mass-ratio-inspirals

### 2014-08-04 res_char.tex
two column layout

\documentclass[aps,prd,amsfonts,amssymb,amsmath,nofootinbib,reprint,showpacs]{revtex4}
\usepackage{graphicx}
\usepackage{color}
%\usepackage{psfrag}
\usepackage{mathrsfs}
\usepackage{dcolumn}
\usepackage{hyperref}

@@

\author{Robert H. Cole}
\email[]{rhc2@ast.cam.ac.uk}
\author{Christopher P.L. Berry}
\email[]{cplb2@ast.cam.ac.uk}
\author{Priscilla Ca\~{n}izares}
\email[]{pcm@ast.cam.ac.uk}
\author{Jonathan R. Gair}
\email[]{jgair@ast.cam.ac.uk}
\affiliation{Institute of Astronomy, Madingley Road, Cambridge, CB3 0HA, United Kingdom}

@@

\begin{abstract}
The inspiral of stellar mass compact objects into supermassive black holes provides a wealth of information about the strong gravitational field regime, via the emission of gravitational waves. In order to detect such signals and extract the desired information, it is necessary to possess a bank of accurate waveform templates. For computational efficiency, adiabatic templates are often used, but these neglect the effect of transient resonances, where the radial and poloidal fundamental frequencies become commensurate. During such resonances, the flux of gravitational waves can be diminished or enhanced, leading to a shift in the compact object trajectory and ultimately the phase of the emitted radiation. An astrophysical population of detectable extreme-mass-ratio-inspirals is studied and it is found that a large proportion will encounter a low-order resonance in the later stages of inspiral. The resulting effect on SNR recovery is small due to the low eccentricity of the systems, but nevertheless at least 20\% of EMRIs may become undetectable if transient resonances aren't accounted for.
\end{abstract}

@@

In the prologue to his classic monograph, Chandrasekhar~\cite{Chandrasekhar1992} celebrates the simplicity of black holes (BHs). The Kerr solution is defined by just two parameters: mass and spin. Despite the baldness of the BH metrics, great intricacies manifest in their properties. This is made evident when a second body is introduced. The two-body problem in general relativity (GR) is well studied. It is of paramount importance for gravitational wave (GW) astronomy, where binary systems are the dominant source of radiation. Correctly modelling the dynamics of these systems is necessary to interpret and extract information from gravitational waveforms.

We have made progress in understanding the general relativistic two-body problem in recent years. Bodies of comparable mass can be studied using numerical relativity. Rapid advances in this field have been made following breakthroughs in 2005~\cite{Pretorius2005,Campanelli2006,Baker2006}. These simulations shall allow us to understand BH--BH mergers. Stellar-mass BH mergers are targets for ground-based GW detectors, such as the currently mid-upgrade LIGO~\cite{Harry2010} and Virgo~\cite{Accadia2011}, and the in-construction Kagra~\cite{Kuroda2010}. Massive BH mergers, expected to be the result of galaxy mergers, are potential sources for a space-borne detector such as eLISA~\cite{Amaro-Seoane2012a}. Systems of unequal masses are more challenging to evolve numerically as they complete a larger number of orbits and it is necessary to resolve two different scales. Calculations can instead be performed perturbatively. The paradigm unequal mass system has a stellar-mass BH orbiting a massive black hole, such as those expected to be found at the centre of most galaxies. These extreme-mass-ratio inspirals (EMRIs) produce GWs that are a promising signal for space-borne detectors~\cite{Amaro-Seoane2007}. Despite being well-studied there still remain open questions.

In the case of extreme-mass-ratio systems, efforts are concentrated on understanding the gravitational self-force~\cite{Barack2009,Poisson2004}. In the test particle limit the smaller body follows an exact geodesic of the BH's spacetime. Including the effects of the smaller body's finite mass, the background spacetime is perturbed. The back-reaction from this deformation alters the small body's orbital trajectory, and can be modelled as a self-force that moves the body from its geodesic. The self-force is commonly divided into two pieces, dissipative and conservative. The former encapsulates the slow decay of the orbital energy and angular momentum, constants of the motion in the test particle limit, through radiation. The latter shifts the orbital phases inducing precession. The dissipative piece is time-asymmetric and has the larger effect on the evolution of the orbital phase; the conservative piece is time-symmetric and has a smaller influence on the phase although this can accumulate over many orbits. Being able to accurately model the influence of the self-force shall allow us to create accurate waveform models.

@@

We also use the resonance time-scale
\begin{equation}
\tau\sub{res} = \left[\frac{2\pi}{\left|\langle\dot{\Omega}(0)\rangle_{q'}\right|}\right]^{1/2}.
\label{eq:t-res}
\end{equation}
Here $\dot{\Omega}(0)$ is the rate of change of $\Omega$ at resonance, which we take to be at $t = 0$. The instantaneous $\dot{\Omega}$ depends upon the orbital phase and oscillates about its mean trend over an orbit (see \figref{wibbles}). We are interested in the averaged behaviour, not the periodic modulations about this, which is why we use the time-average $\langle\dot{\Omega}\rangle_{q'}$; here we use $q'$ to represent a phase that varies over an orbit with period of order $T$. Close to resonance, $\Omega(t)$ is well approximated by a first-order Taylor expansion, decreasing linearly with time; hence we make the approximation
\begin{equation}
\left|\langle\dot{\Omega}\rangle_{q'}\right| \simeq \left|\frac{\Omega(t)}{t}\right|.
\end{equation}
The resonant time-scale should give an indication of the time over which we expect the effects of the resonance to be felt~\cite{Bosley1992}. Consider the phase of the Mino time Fourier expansion on resonance; neglecting the constant, the resonant Fourier component has form
\begin{equation}
\varphi_{n_r,\,-n_\theta} \simeq \left(n_r\Upsilon_r - n_\theta\Upsilon_\theta\right)\lambda + \left(n_r\dot{\Upsilon_r} - n_\theta\dot{\Upsilon}_\theta\right)\lambda^2 + \ldots
\end{equation}
Typically, the first term is non-zero and this gives the familiar oscillation. On resonance, it is zero, leaving the next order term to govern the behaviour~\cite{Flanagan2012}. Only once we have moved far enough away from resonance for the first term to dominate the second do we recapture the non-resonant behaviour. The first term (translating from Mino time to coordinate time) sets $\tau\sub{pres}$, the second sets $\tau\sub{res}$.

@@

Once we have established a waveform generation procedure, we can define a natural inner product between two waveforms, $s_A(t)$ and $h_A(t)$, according to
\begin{equation}
\label{eq:innerprod}
\overlap{s}{h} = 2 \sum_{A=+,\times}\intd{0}{\infty}{\frac{\tilde{s}_A(f)\tilde{h}_A^*(f)+\tilde{s}_A^*(f)\tilde{h}_A(f)}{S_n(f)}}{f},
\end{equation}
where $\tilde{s}_A(f)$ represents the Fourier transform of $s_A(t)$ and similarly for $\tilde{h}_A(f)$, $A = \{+,\times\}$ denotes the plus- or cross-polarisation state of the GW and $S_n(f)$ is the one-sided noise power spectral density (PSD). We are interested in the detectability of EMRIs using future space-based GW detectors and so we use a PSD appropriate for describing the eLISA mission concept. The analytic approximation we use is~\cite{Amaro-Seoane2012a}
\begin{equation}
S_n(f) = \frac{20}{3}\frac{4 S_{acc}(f)/(2\pi f)^4 + S_{sn}(f) + S_{omn}(f)}{L^2} \left(1+\left(\frac{f}{0.41 c/2L}\right)\right)^2,
\end{equation}
where $L = 1\;\mathrm{Mkm}$ is the length of the interferometer arms, $c$ is the speed of light and the components of noise separate into: the noise due to residual test mass acceleration $S_{acc}(f) = 2.13\times10^{-29}(1+10^{-4}\mathrm{Hz}/f)\;\mathrm{m^2\,s^{-4}\,Hz^{-1}}$; the shot noise $S_{sn} = 5.25\times10^{-23}\;\mathrm{m^2\,Hz^{-1}}$; and all other measurement noises $S_{omn} = 6.28\times10^{-23}\;\mathrm{m^2\,Hz^{-1}}$. We show a plot of this PSD against frequency in \figref{noise-curves}, along with the corresponding curve for the LISA detector for comparison. The effect of white dwarf (WD) confusion noise, which is important for LISA (as shown in the figure), doesn't significantly impact the eLISA mission and so we neglect it.

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=0.92\textwidth]{Fig_good_traj}
\caption{\label{fig:good-traj}The evolution of the orbital parameters $E$ (left), $L_z$ (center) and $Q$ (right) under the full (solid line) and adiabatic (dashed) models for an illustrative EMRI system that does not encounter any significant resonances. The inset plots show the behaviour on short timescales, where the fast orbital oscillations can be seen.}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=0.92\textwidth]{Fig_good_waveform}
\caption{\label{fig:good-waveform}The plus-polarised waveform for an illustrative EMRI system with $p=7.5$, shown for two short segments at the start and end of a $2$ year evolution. The full and adiabatic models are both shown, but are indistinguishable by eye.}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=0.92\textwidth]{Fig_dephased_waveform}
\caption{\label{fig:dephased-waveform}The plus-polarised waveform for an illustrative EMRI system with $p=7.85$, shown for two short segments at the start and end of a $2$ year evolution. The full and adiabatic models are both shown; there is a significant dephasing between them by the end of the evolution.}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{Fig_res_diff_traj}
\caption{\label{fig:res-diff-traj}The differences in orbital parameters $E$ (left), $L_z$ (center) and $Q$ (right) between each evolution scheme and the adiabatic model that matches at the start. The solid line shows the full evolution and the dashed lines show the different adiabatic evolutions, which match the full evolution at different times throughout the inspiral (numbers in parentheses give the overlap with the full evolution): at the start (0.202), at the 2:3 resonance (0.458), after the 2:3 resonance (0.259) and at the end (0.677).}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{Fig_overlap_vs_resjump_ill}
\caption{\label{fig:overlap-vs-resjump-ill}The overlap \eqref{eq:overlap} between the full and adiabatic evolutions, maximised over the family of generated adiabatics, as a function of the resonance jumps in $E$ (left), $L_z$ (middle), and $Q$ (right). Each dot represents a system with our illustrative parameters and a random choice of the initial radial phase. The red line indicates the expected overlap assuming that the waveforms match exactly post-resonance, but have zero overlap pre-resonance.}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{Fig_EMRIpar_dists}
\caption{\label{fig:EMRIpar-dists}The parameter distributions for our detectable EMRI systems, generated using $50,000$ realisations. The (darker) blue bars represent the final distribution, including an SNR cut of 10. The (lighter) pink bars stacked on top indicate the full distribution before the SNR cut.}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=0.92\textwidth]{Fig_pop_SNR_vs_jump}
\caption{\label{fig:pop-SNR-vs-jump}The fraction of the full SNR recovered using our family of adiabatic models, as a function of the percentage resonance jump in $E$. The histograms show the marginalised distributions over each parameter.}
\end{figure}

>>>>

\documentclass[aps,prd,amsfonts,amssymb,amsmath,nofootinbib,reprint,showpacs,superscriptaddress,twocolumn]{revtex4}
\usepackage{graphicx}
\usepackage{color}
%\usepackage{psfrag}
\usepackage{mathrsfs}
\usepackage{dcolumn}
\usepackage{hyperref}

@@

\author{Robert H. Cole}
\email[]{rhc26@ast.cam.ac.uk}
\affiliation{Institute of Astronomy, Madingley Road, Cambridge, CB3 0HA, United Kingdom}
\author{Christopher P.L. Berry}
\email[]{cplb@star.sr.bham.ac.uk}
\affiliation{School of Physics and Astronomy, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom}
\author{Priscilla Ca\~{n}izares}
\email[]{pcm@ast.cam.ac.uk}
\author{Jonathan R. Gair}
\email[]{jgair@ast.cam.ac.uk}
\affiliation{Institute of Astronomy, Madingley Road, Cambridge, CB3 0HA, United Kingdom}

@@

\begin{abstract}
The inspiral of stellar-mass compact objects into supermassive black holes provides a wealth of information about the strong gravitational-field regime via the emission of gravitational waves. In order to detect such signals and extract the desired information, it is necessary to possess a bank of accurate waveform templates. For computational efficiency, adiabatic templates are often used, but these neglect the effect of transient resonances, where the radial and poloidal fundamental frequencies become commensurate. During such resonances, the flux of gravitational waves can be diminished or enhanced, leading to a shift in the compact-object trajectory and, ultimately, the phase of the emitted radiation. In this work, an astrophysical population of detectable extreme-mass-ratio-inspirals is studied and it is found that a large proportion shall encounter a low-order resonance in the later stages of inspiral. The resulting effect on signal-to-noise recovery is small due to the low eccentricity of the studied population, but nevertheless at least one fifth of the systems may become undetectable if transient resonances aren't accounted for.
\end{abstract}

@@

In the prologue to his classic monograph, Chandrasekhar~\cite{Chandrasekhar1992} celebrates the simplicity of black holes (BHs). The Kerr solution is defined by just two parameters: mass and spin. Despite the baldness of the BH metrics, great intricacies manifest in their properties. This is made evident when a second body is introduced. The two-body problem in general relativity (GR) is well studied. It is of paramount importance for gravitational-wave (GW) astronomy, where binary systems are the dominant source of radiation. Correctly modelling the dynamics of these systems is necessary to interpret and extract information from gravitational waveforms.

We have made progress in understanding the general relativistic two-body problem in recent years. Bodies of comparable mass can be studied using numerical relativity. Rapid advances in this field have been made following breakthroughs in 2005~\cite{Pretorius2005,Campanelli2006,Baker2006}. These simulations shall allow us to understand BH--BH mergers. Stellar-mass BH mergers are targets for ground-based GW detectors, such as the currently mid-upgrade LIGO~\cite{Harry2010} and Virgo~\cite{Accadia2011}, and the in-construction KAGRA~\cite{Kuroda2010}. Massive BH mergers, expected to be the result of galaxy mergers, are potential sources for a space-borne detector such as eLISA~\cite{Amaro-Seoane2012a}. Systems of unequal masses are more challenging to evolve numerically as they complete a larger number of orbits and it is necessary to resolve two different scales. Calculations can instead be performed perturbatively. The paradigm unequal-mass system has a stellar-mass BH orbiting a massive black hole, such as those expected to be found at the centre of most galaxies. These extreme-mass-ratio inspirals (EMRIs) produce GWs that are a promising signal for space-borne detectors~\cite{Amaro-Seoane2007}. Despite being well-studied there still remain open questions.

In the case of extreme-mass-ratio systems, efforts are concentrated on understanding the gravitational self-force~\cite{Barack2009,Poisson2004}. In the test particle limit, the smaller body follows an exact geodesic of the BH's spacetime. Including the effects of the smaller body's finite mass, the background spacetime is perturbed. The back-reaction from this deformation alters the small body's orbital trajectory, and can be modelled as a self-force that moves the body from its geodesic. The self-force is commonly divided into two pieces, dissipative and conservative. The former encapsulates the slow decay of the orbital energy and angular momentum, constants of the motion in the test particle limit, through radiation. The latter shifts the orbital phases inducing precession. The dissipative piece is time-asymmetric and has the larger effect on the evolution of the orbital phase; the conservative piece is time-symmetric and has a smaller influence on the phase although this can accumulate over many orbits. Being able to accurately model the influence of the self-force shall allow us to create reliable waveform models.

@@

We also use the resonance time-scale
\begin{equation}
\tau\sub{res} = \left[\frac{2\pi}{\left|\langle\dot{\Omega}(0)\rangle_{q'}\right|}\right]^{1/2}.
\label{eq:t-res}
\end{equation}
Here $\dot{\Omega}(0)$ is the rate of change of $\Omega$ at resonance, which we take to be at $t = 0$. The instantaneous $\dot{\Omega}$ depends upon the orbital phase and oscillates about its mean trend over an orbit. We are interested in the averaged behaviour, not the periodic modulations about this, which is why we use the time-average $\langle\dot{\Omega}\rangle_{q'}$; here we use $q'$ to represent a phase that varies over an orbit with period of order $T$. Close to resonance, $\Omega(t)$ is well approximated by a first-order Taylor expansion, decreasing linearly with time; hence we make the approximation
\begin{equation}
\left|\langle\dot{\Omega}\rangle_{q'}\right| \simeq \left|\frac{\Omega(t)}{t}\right|.
\end{equation}
The resonant time-scale should give an indication of the time over which we expect the effects of the resonance to be felt~\cite{Bosley1992}. Consider the phase of the Mino time Fourier expansion on resonance; neglecting the constant, the resonant Fourier component has form
\begin{equation}
\varphi_{n_r,\,-n_\theta} \simeq \left(n_r\Upsilon_r - n_\theta\Upsilon_\theta\right)\lambda + \left(n_r\dot{\Upsilon_r} - n_\theta\dot{\Upsilon}_\theta\right)\lambda^2 + \ldots
\end{equation}
Typically, the first term is non-zero and this gives the familiar oscillation. On resonance, it is zero, leaving the next order term to govern the behaviour~\cite{Flanagan2012}. Only once we have moved far enough away from resonance for the first term to dominate the second do we recapture the non-resonant behaviour. The first term (translating from Mino time to coordinate time) sets $\tau\sub{pres}$, the second sets $\tau\sub{res}$.

@@

Once we have established a waveform generation procedure, we can define a natural inner product between two waveforms, $s_A(t)$ and $h_A(t)$, according to
\begin{equation}
\label{eq:innerprod}
\overlap{s}{h} = 2 \sum_{A=+,\times}\intd{0}{\infty}{\frac{\tilde{s}_A(f)\tilde{h}_A^*(f)+\tilde{s}_A^*(f)\tilde{h}_A(f)}{S_n(f)}}{f},
\end{equation}
where $\tilde{s}_A(f)$ represents the Fourier transform of $s_A(t)$ and similarly for $\tilde{h}_A(f)$, $A = \{+,\times\}$ denotes the plus- or cross-polarisation state of the GW and $S_n(f)$ is the one-sided noise power spectral density (PSD). We are interested in the detectability of EMRIs using future space-based GW detectors and so we use a PSD appropriate for describing the eLISA mission concept. The analytic approximation we use is~\cite{Amaro-Seoane2012a}
\begin{widetext}\begin{equation}
S_n(f) = \frac{20}{3}\frac{4 S_{acc}(f)/(2\pi f)^4 + S_{sn}(f) + S_{omn}(f)}{L^2} \left(1+\left(\frac{f}{0.41 c/2L}\right)\right)^2,
\end{equation}\end{widetext}
where $L = 1\;\mathrm{Mkm}$ is the length of the interferometer arms, $c$ is the speed of light and the components of noise separate into: the noise due to residual test mass acceleration $S_{acc}(f) = 2.13\times10^{-29}(1+10^{-4}\mathrm{Hz}/f)\;\mathrm{m^2\,s^{-4}\,Hz^{-1}}$; the shot noise $S_{sn} = 5.25\times10^{-23}\;\mathrm{m^2\,Hz^{-1}}$; and all other measurement noises $S_{omn} = 6.28\times10^{-23}\;\mathrm{m^2\,Hz^{-1}}$. We show a plot of this PSD against frequency in \figref{noise-curves}, along with the corresponding curve for the LISA detector for comparison. The effect of white dwarf (WD) confusion noise, which is important for LISA (as shown in the figure), doesn't significantly impact the eLISA mission and so we neglect it.

@@

\begin{figure*}[htbp]
\centering
\includegraphics[width=0.92\textwidth]{Fig_good_traj}
\caption{\label{fig:good-traj}The evolution of the orbital parameters $E$ (left), $L_z$ (center) and $Q$ (right) under the full (solid line) and adiabatic (dashed) models for an illustrative EMRI system that does not encounter any significant resonances. The inset plots show the behaviour on short timescales, where the fast orbital oscillations can be seen.}
\end{figure*}

@@

\begin{figure*}[htbp]
\centering
\includegraphics[width=0.92\textwidth]{Fig_good_waveform}
\caption{\label{fig:good-waveform}The plus-polarised waveform for an illustrative EMRI system with $p=7.5$, shown for two short segments at the start and end of a $2$ year evolution. The full and adiabatic models are both shown, but are indistinguishable by eye.}
\end{figure*}

@@

\begin{figure*}[htbp]
\centering
\includegraphics[width=0.92\textwidth]{Fig_dephased_waveform}
\caption{\label{fig:dephased-waveform}The plus-polarised waveform for an illustrative EMRI system with $p=7.85$, shown for two short segments at the start and end of a $2$ year evolution. The full and adiabatic models are both shown; there is a significant dephasing between them by the end of the evolution.}
\end{figure*}

@@

\begin{figure*}[htbp]
\centering
\includegraphics[width=\textwidth]{Fig_res_diff_traj}
\caption{\label{fig:res-diff-traj}The differences in orbital parameters $E$ (left), $L_z$ (center) and $Q$ (right) between each evolution scheme and the adiabatic model that matches at the start. The solid line shows the full evolution and the dashed lines show the different adiabatic evolutions, which match the full evolution at different times throughout the inspiral (numbers in parentheses give the overlap with the full evolution): at the start (0.202), at the 2:3 resonance (0.458), after the 2:3 resonance (0.259) and at the end (0.677).}
\end{figure*}

@@

\begin{figure*}[htbp]
\centering
\includegraphics[width=\textwidth]{Fig_overlap_vs_resjump_ill}
\caption{\label{fig:overlap-vs-resjump-ill}The overlap \eqref{eq:overlap} between the full and adiabatic evolutions, maximised over the family of generated adiabatics, as a function of the resonance jumps in $E$ (left), $L_z$ (middle), and $Q$ (right). Each dot represents a system with our illustrative parameters and a random choice of the initial radial phase. The red line indicates the expected overlap assuming that the waveforms match exactly post-resonance, but have zero overlap pre-resonance.}
\end{figure*}

@@

\begin{figure*}[htbp]
\centering
\includegraphics[width=\textwidth]{Fig_EMRIpar_dists}
\caption{\label{fig:EMRIpar-dists}The parameter distributions for our detectable EMRI systems, generated using $50,000$ realisations. The (darker) blue bars represent the final distribution, including an SNR cut of 10. The (lighter) pink bars stacked on top indicate the full distribution before the SNR cut.}
\end{figure*}

@@

\begin{figure*}[htbp]
\centering
\includegraphics[width=0.92\textwidth]{Fig_pop_SNR_vs_jump}
\caption{\label{fig:pop-SNR-vs-jump}The fraction of the full SNR recovered using our family of adiabatic models, as a function of the percentage resonance jump in $E$. The histograms show the marginalised distributions over each parameter.}
\end{figure*}
